Data Descriptor for Nature Scientific Data

### 2024-11-29 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-updates

However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities in our societies [@chang2020].
In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].

@@

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their relative level of socioeconomic deprivation.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.

@@

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days.
The extent of this decrease tends to be larger for high-density areas.
Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%.
However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes.
Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions.
In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
Notably, mobility levels in areas characterised by higher population density still remain below the baseline.
Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
Overall, the observed patterns by population density category are consistent with those reported in [@RoweCCA24] for Argentina and Chile, where data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.

@@

![Figures](figures/Fig-recovery-density.png){#fig-recoverydensity}

@@

![Figures RDI](figures/Fig-recovery-rdi.png){#fig-recoveryrdi}

@@

Figures @fig-redensity and @fig-rerdi present, for each country, the random variations in intercept and slope estimated by Models 3, 4, and 5, providing insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
Focusing first on @fig-redensity, we observe that

@@

## Focus on cities {#sec-cities}

![FUAS netflows](figures/Fig-fuas-netflows.pdf){#fig-cities}

@@

@@

## Meta-Facebook data

@@

## WorldPpop population data

@@

## Socioeconomic deprivation data

@@

# Methods

## Classification of tiles according to level of urbanisation and socioeconomic deprivation

The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.

@@

Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
For example, Figure @Fig1 shows, for each country, the proportion of the population across the various analysis categories.
These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
This issue is revisited in the Discussion section.

![Classification of spatial units into categories by population density and by relative deprivation index. Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#Fig1}

## Processing Facebook data

@@

## Trend analysis

@@

# Code availability {.unnumbered}

@@

::: {#refs}
:::

>>>>

However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities in our societies [@chang2020].
In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].
At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].

@@

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their relative level of socioeconomic deprivation.
More details for the classification method are provided in the Methods section.
The two criteria for classification are chosen due to their relevance for *`@Francisco, can you help here?`*.
The analysis of mobility is done using Facebook movement data that has been processed according to our methodology.

@@

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days.
The extent of this decrease tends to be larger for high-density areas.
Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage change exceeding 100%.
However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage change.
Nonetheless, these relative changes illustrate the extent of the impacts experienced by these sparsely populated regions.
In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
Notably, mobility levels in areas characterised by higher population density still remain below the baseline.
Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
Overall, the observed patterns by population density category are consistent with those reported in [@RoweCCA24] for Argentina and Chile, where data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.

@@

![Evolution of number of flows by population density category of origin for Argentina, Chile and Colombia, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png){#fig-recoverydensity}

@@

![Evolution of number of flows by population density category of origin, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png)

@@

Figures @fig-redensity and @fig-rerdi present, for each country, the random variations in intercept and slope estimated by Models 3, 4, and 5, providing insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
Focusing first on @fig-redensity, we observe that...
*`@Francisco to write this interpretation`*

@@

## Focus on cities - *`@Francisco @Miguel, we need a better title here - any ideas?`* {#sec-cities}

Turning our focus to cities, we present a city-centric analysis of the evolution of movements.
Specifically, we examine movements between urban centers and the rest of spatial units within each country, categorising these spatial units by population density to account for variability across the urban hierarchy.
Urban centers are defined as the most densely populated spatial unit within each Functional Urban Area (FUA).
The boundaries of these Functional Urban Areas are derived from the European Commission's Human Settlement Layer (GHSL) project [@ceu.jrc.2019] .

For each country, we analyze inflows to all FUA centers, considering two types of movements: those originating from any spatial unit (first column of @fig-fuasflows) and those from spatial units within a 100 km radius of the FUA centers (second column of @fig-fuasflows).
Additionally, focusing only on urban centres corresponding to the capital FUAs -Buenos Aires, Santiago and Bogotá- we also look at inflows from any spatial unit and from those situated within a 100 km radius (columns third and fourth columns in @fig-fuasflows).
The analysis is similarly repeated for outflows from urban centers to other spatial units.

*`@Miguel to write this section. What can we learn about the figure? The target journal is nature cities, so we need to talk about cities...`*

![Evolution of number of inflows (and outflows) to (from) functional urban area centres, by population desnity category of origin (destiantion) for Argentina, Chile and Colombia, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-fuas-flows.png){#fig-fuasflows}

@@

*`@Miguel to revisit this section - the text here was written a long time ago, before the analysis had taken the direction it's taken. Needs to be rewritten. Focus on trajectories of recovery. Focus on differences across population groups. Talk about remaining un-addressed biases in the data. With our methodology what we achieve is 1) data imputation for missing counts in regions that have low counts due to a variety of reasons -discuss- 2) data adjustment for fluctuations in daily number of users. However, our method does not explicitely adjust underrepresentation of specific population groups. This remains as future work and we should acknowledge the importance of this to make research based on digital traces more relevant for policy.`*

@@

## Meta-Facebook data {#sec-datameta}

@@

## WorldPpop population data {#sec-dataworldpop}

@@

## Socioeconomic deprivation data {#sec-datardi}

@@

# Methods {#sec-methods}

## Classification of tiles according to level of urbanisation and socioeconomic deprivation {#sec-methodsclassify}

The geographic distribution of categories for each of the countries is displayed in @fig-classification for Argentina, Chile and Colombia respectively.

@@

Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
For example, Figure @fig-classification shows, for each country, the proportion of the population across the various analysis categories.
These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
This issue is revisited in the Discussion section.

![Classification of spatial units into categories by population density and by relative deprivation index. Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#fig-classification}

## Processing Facebook data {#sec-methodsprocess}

@@

## Trend analysis {#sec-methodstrend}

@@

# Code availability

@@
